Assessing latent variables models by studying the moment estimator gap

### 2021-02-08 Draft.tex
ppca (incomplete)
Completed ppca, tiny bits of article update

\section{Intro}

@@

In the next section we establish the family of latent variable model suited for this metric. In section 3 we discuss the moments estimators we use before building the proposed evaluation metric in section 4. After we discuss related work in section 5 we present simulations and experiments in section 6. Finally, section 7 propose a wide range of future application of the ideas presented in this article such as the possibility of training models using the proposed metric.

\section{Latent variable generative models}

The diagnostic tool we propose is useful to assess the quality of the fit for a family of unsupervised latent variable models. In this section we define this family.

@@

\end{tikzpicture}
  \caption{Graphical representation of latent variables models with joint distribution $p(\mathbf{x},\mathbf{z}) = p(\mathbf{z})p(\mathbf{x}|z)$.}
  \label{genny}
\end{figure}%
These models are \textit{generative models} because learning $p(\x,\z)= p(\z)p(\x|z)$ allows us to generated new samples of $\x$ using ancestral sampling. What makes this model probabilistic is that the mapping from $\z$
 to $\x$ is not a deterministic function $f: \mathcal{Z} \rightarrow \mathcal{X}$ but instead a \textit{probabilistic mapping} from $\mathcal{Z}$ to the parameter space $\Theta$, where $\Theta$ is the parameter space of $p_\theta(\x|z)$; $\theta \in \Theta$.

@@

When training or fitting such models we are training the function $f:\mathcal{Z} \rightarrow \Theta$ in ways to maximize the likelihood of the data set $S$ under the model of equation \ref{LVM}. This mapping $f$ explains the effect of $\z$ on $\x$ ans is at the centre of latent variable models. Learning this function $f$ is the main challenge of optimizing latent variable model and the Expectation-Maximization algorithm (EM) or variational inference are common strategies to fit learn such function.

@@

Usually $p_\theta(\x|z)$ is a simple parametric distribution and the latent variable is in charge of increasing the complexity of $p_\theta(\x)$. Additionally, the function $f$ can take many forms, from simple linear combination to neural network functions. We will use $f(\z)$ interchangeably with $f$ but also sometimes the \textit{distribution parameters} it outputs directly. For instance, assuming $p_\theta(\x|z) = \text{Poisson}(\lambda)$ then $f: \mathcal{Z} \rightarrow \mathbb{R}^+$ and we might use $f(\z)$ and $\lambda(\z)$ interchangeably.

@@

The GMM is a member of the family latent variable generative models we just described. For a GMM with $K$-components we define $\z$ as a k-class categorical variable, $\z \in \{1,...,K\} = \mathcal{Z}$, $p(\z)$ is a categorical distribution where we define $\pi_j$ as $p(\z = j)$ with $\sum_{j=1}^K \pi_j =1$. Finally, setting $p(\x|z=j) = N(\mu_j,\sigma_j)$ leads to a GMM:
\begin{align}
p(\x) = \sum_{j=1}^K \pi_j p(\x|z=j) 
\label{LVM}
\end{align}
In this situation, $f$ map the latent variable to a pair of \textit{distribution parameters}, $\mu$ and $\sigma$, $f: \{1,...,K\} \rightarrow \mathbb{R}\times\mathbb{R}^+$. Similarly

\section{Moment estimators}

@@

\begin{align}
\Vx(X) = \Ez[\Vx(X|Z)] + \Vz[\Ex(X|Z)]
\end{align}

@@

\begin{align}
\Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
&\approx \frac{1}{m} \sum_{i=1}^m ( \text{diag}(\text{MLP}_2(z_i)) + \text{MLP}_1(z_i)^T\text{MLP}_1(z_i))
\label{rhs}
\end{align}

where $z_i \sim p_\theta(z)$, MLP$_1$ is the multilayer perceptron that return $\Ex(X|Z)$ and MLP$_2$ is the multilayer perceptron that return $\Vx(X|Z)$. This is the forward model estimates (FME). Notice that this estimator does not require we sample from $p_\theta(x|z)$ and uses directly the \textit{parameters} of this distribution instead. It is very fast to sample a large amount of $z's$ and additionally, because this is a rather simple Monte Carlo sample, we can compute its estimated standard error ...

@@

\begin{align}
\Ex[X] = \Ez[\Ex(X|Z)] 
\label{rhs}
\end{align}

@@

\section{MEGA (Moment Estimators GAp)}

@@

\section{Related Works}

\section{Experiments and results}

\section{Future work}
\subsection{Extension to a wider class of probabilistic graphical models}
\subsection{Training by minimizing MEGA}

\section{Conclusion}

>>>>

\section{Introduction} \label{intr}

@@

In the next section we establish the family of latent variable model suited for this metric. After we discuss related work in section \ref{rewo}, we discuss the moments estimators used in section \ref{moes}.  In section \ref{mega}, we present our metric. We show how this metric perform with simulations in section \ref{expe}. Finally, section \ref{fuwo} propose a wide range of future application of the ideas presented in this article such as the possibility of training models using the proposed metric.

\section{Latent variable generative models}\label{lvgm}

The diagnostic tool we propose is useful to assess the quality of the fit for a family of unsupervised latent variable models. In this section we define this family, the Latent variable generative models (LVGMs).

@@

\end{tikzpicture}
  \caption{Graphical representation of latent variables models with joint distribution $p(\mathbf{x},\mathbf{z}) = p(\mathbf{z})p(\mathbf{x}|z)$.}
  \label{genny}
\end{figure}%
These models are \textit{generative models} because learning $p(\x,\z)= p(\z)p(\x|z)$ allows us to generated new samples of $\x$ using ancestral sampling. What makes this model probabilistic is that the mapping from $\z$
 to $\x$ is not a deterministic function $f: \mathcal{Z} \rightarrow \mathcal{X}$ but instead a \textit{probabilistic mapping} from $\mathcal{Z}$ to the parameter space $\Theta$, where $\Theta$ is the parameter space of $p_\theta(\x|z)$; $\theta \in \Theta$. We will call $p_\theta(\x|z)$ the emission distribution or observation distribution interchangeably.

@@

When training or fitting such models we are training the function $f:\mathcal{Z} \rightarrow \Theta$ in ways to maximize the likelihood of the data set $S$ under the model of equation \ref{LVM}. This mapping $f$ explains the effect of $\z$ on $\x$ ans is at the centre of latent variable models. Learning this function $f$ is the main challenge of optimizing latent variable model and the Expectation-Maximization algorithm (EM) or variational inference are common strategies to fit learn such function. In most cases, $p(\z)$ is assumed to be known and fixed but in some cases the parameters of $p(\z)$ are estimated as well.

@@

Usually $p_\theta(\x|z)$ is a simple parametric distribution and the latent variable is in charge of increasing the complexity of $p_\theta(\x)$. Additionally, the function $f$ can take many forms, from simple linear combination to neural network functions. We will use $f(\z)$ interchangeably with $f$ but also sometimes the \textit{distribution parameters} it outputs directly. For the majority of emission distribution we consider, there exist a simple mapping from the parameters of the distribution to its expectation and its variance.

For instance, assuming the emission distribution is Poisson: $p_\theta(\x|z) = \text{Poisson}(\lambda)$ then $f: \mathcal{Z} \rightarrow \mathbb{R}^+$ and we might use $f(\z)$ and $\lambda(\z)$ interchangeably. Additionally, $\Ex[\x|\z] = \lambda(\z)$ and $ \Vx[\x|\z] = \lambda(\z)$.

The Probabilistic Principal Component Analysis (pPCA) \cite{tipping99,Bishop07} is a member of the LVGM family we just described where $p(\z)$ is assumed to be a normal distribution $N(0,I)$. We also assume the emission distribution to be Normal: $p(\x|\z) = N(W\z+\mu,\sigma^2I)$. In this formulation, we see that $f: \mathbb{R}_M \rightarrow \mathbb{R}_D$ is a linear function that maps the latent variable $\z$ to $\Ex[\x|\z]$. Outside of estimating $W$ and $\mu$ as part of $f$ the model also estimate $\sigma^2$ though it is not a function of $z$. However it is function of $D$ and $M$, the dimension of $\mathcal{X}$ and $\mathcal{Z}$.

The parameters of pPCA can be fit analytically as the solution of a direct maximization of the likelihood or with the EM algorithm.

@@

For a GMM with $K$-components we define $\z$ as a k-class categorical variable, $\z \in \{1,...,K\} = \mathcal{Z}$, $p(\z)$ is a categorical distribution where we define $\pi_j$ as $p(\z = j)$ with $\sum_{j=1}^K \pi_j =1$. Finally, setting $p(\x|z=j) = N(\mu_j,\sigma_j)$ leads to a GMM:
\begin{align}
p(\x) = \sum_{j=1}^K \pi_j p(\x|z=j) 
\label{LVM}
\end{align}
In this situation, $f$ map the latent variable to a pair of \textit{distribution parameters}, $\mu$ and $\sigma$, $f: \{1,...,K\} \rightarrow \mathbb{R}\times\mathbb{R}^+$. In this particular cases  $\Ex[\x|\z] = f_1(\z)$ where $f_1$ is the \textit{first} output of $f(\z)$ and $\Vx[\x|\z] = f_2(\z)$.

The GMM is a special case of LVGM where we also estimate the parameters $\{\pi_j : j \in 1,...K\}$ of $p_(\z)$. A GMM is usually train using the EM algorithm.

\section{Related Works} \label{rewo}

We propose a metric to evaluate to goodness-of-fit of the family of latent variable define in section \ref{lvgm} and based on the literature there is no good metric for such thing at the moment.

In the literature the likelihood of the data $S$ under the fitted model is used CITE, however,it has been shown CITE that sometimes a model can reach very high-likelihood even for terrible fit. For instance, Bishop et al. \cite{Bishop07} demonstrate that for

\section{Moment estimators} \label{moes}

@@

\begin{align}
\Vx(\x) = \Ez[\Vx(\x|\z)] + \Vz[\Ex(\x|\z)]
\end{align}

@@

\begin{align}
\Ez[\Vx(X|Z) + \Ex(X|Z)^2] &= \int_z (\Vx(X|Z=z)+\Ex(X|Z=z)^2)p(z) dz \\ 
&\approx \frac{1}{m} \sum_{i=1}^m \left[\Vx(\x|\z=z_i) + \Ex(\x|\z=z_i)^T\Ex(\x|\z=z_i)\right]
\label{rhs}
\end{align}

where $z_i \sim p_(\z)$, and $\Ex(\x|\z=z_i)$ and  $\Vx(\x|\z=z_i)$ are expressed as functions of $f(z_i)$. Thus this estimator relies on all the components of the fitted LVGM: $p(\z)$ and $f(\z)$. This is the forward model estimates (FME). Notice that this estimator does not require we sample from $p_\theta(x|z)$ and uses directly the estimated parameters of the emission distribution. It is very fast to sample a large amount of $z's$ and additionally, because this is a rather simple Monte Carlo sample, we can compute its estimated standard error ...

@@

\begin{align}
\Ex[X] = \Ez[\Ex(X|Z)] 
\label{1stm}
\end{align}

@@

\section{MEGA: Moment Estimators GAp} \label{mega}

@@

\section{Experiments and results} \label{expe}

\section{Future work} \label{fuwo}
\subsection{Extension to a wider class of probabilistic graphical models}
\subsection{Training by minimizing MEGA}

\section{Conclusion} \label{conc}
